Supplementary material

### 2021-11-01 analysis/paper/paper.Rmd
Stepping through each site

# Introduction
The relative sea-level fall that characterises most areas of Fennoscandia is fundamental to its archaeology, as archaeological phenomena that were originally shore-bound can be dated with reference to relative sea-level change . This follows from the fact that if the phenomena were situated on or close to the contemporary shoreline, a reconstruction of the trajectory of shoreline displacement can be used to date the original occupation of the sites based on their elevation relative to the present day sea level.

@@

The close association between Norwegian Stone Age sites located in coastal regions and prehistoric shorelines was firmly established by the geologist W.C. Brøgger and his son, the archaeologist A.W. Brøgger, in their respective theses of 1905 (see also @maccurdy1906). One of the major implications of this finding is thus that relative sea-level change can be used for the dating of the sites. The method of shoreline dating has consequently been fundamental to Stone Age research in Norway ever since. The method is used both independently and to compliment the dating of sites where other sources of temporal data such as typological indicators or radiometric dates are either limited or not available. Given the coarse and fuzzy resolution of established typological frameworks, the vast amount of surveyed sites that only contain generic lithic debitage that could hail from any part of the period, and as the conditions for the preservation of organic material is typically poor in Norway, dating by means of shoreline displacement is often the only and most precise and accurate method by which one can hope to date the sites. Dating by means of shoreline displacement is consequently fundamental to our understanding of the Norwegian Stone Age. However, this is both because shoreline dating is fundamental to the temporal framework on which our understanding of the period is based, but also because the method is only applicable so long as the societies in question have continuously settled on or close to the contemporary shore-line. Consequently, adherence or deviation from this pattern has major implications for the socio-economic foundations of the societies in question.

Despite the fundamental role of shoreline-dating in Norway, there have only been conducted relatively small scale investigations into the applicability of dating by means of shore-line displacement, and the relationship has only been evaluated using what could be termed relatively coarse methods. The aim of this paper is to provide a systematic and comprehensive review of the degree to which radiocarbon dates correspond with the dates informed by our current knowledge of shoreline displacement in a larger area of south-eastern Norway, using a more refined methodological approach. The goal is to quantify the degree to which the assumption of shore-bound settlement holds through the Stone Age. This can be used both to inform the certainty with which one can depend on dating by means of shoreline displacement as well as its precision. The investigation can also help identify societal shifts, if there is chronological variation in the degree to which the sites are shore-bound, and any such variation can potentially also have implications for how archaeological investigations should be directed at different altitudes in the area of analysis. As presented in more detail below, this problem involves the combined evaluation of three major components. One is the questions of when the sites were in use, the second pertains to the reconstruction of the contemporaneous sea-level, and the third follows from the fact that the relation between site and shore-line is inherently a spatial question. A Monte Carlo simulation-based approach was adopted to integrate all three dimensions in the analysis. The approach is only used to inform the question of the distance between sites and shorelines here, but this method can also be extended to a wide range of use-cases where one needs to evaluate and visualise the relationship between archaeological phenomena, the prehistoric shore-line, and the uncertainty inherent in this reconstruction.

@@

# Methods
The goal is effectively to evaluate the spatial relationship between two phenomena, each associated with their own range of temporal uncertainty. First, a single year was drawn from the posterior desnity estimate. This then corresponds to a possible elevation range in the shoreline displacement curve, adjusted to the site location. An elevation value was drawn uniformly from this range, using intervals of 5 cm. The sea-level was then raised to this elevation on the digital terrain model, and the vertical, horisontal and topographic distance between site and shore-line was measured. This process was repeated 1000 times for each phase for each site.

>>>>

# Introduction
The post-glacial relative sea-level fall that characterises most areas of Fennoscandia is fundamental to its archaeology. This follows not only from the dramatic changes to the landscape that this process represent, but also from the fact that if archaeological phenomena were situated on or close to the contemporary shoreline when they were in use, a reconstruction of the trajectory of shoreline displacement can be used to date these phenomena based on their altitude relative to the present day sea-level.

The close association between Norwegian Stone Age sites located in coastal regions and prehistoric shorelines was established by the geologist W.C. Brøgger and his son, the archaeologist A.W. Brøgger, in their respective theses of 1905 (see also @maccurdy1906). The method of shoreline dating has been fundamental to Stone Age research in Norway ever since, and is used both independently and to compliment the dating of sites where other sources of temporal data such as typological indicators or radiometric dates are either limited or not available. Given the coarse and fuzzy resolution of established typological frameworks, the vast amount of surveyed sites that only contain generic lithic debitage that could hail from any part of the period, and as the conditions for the preservation of organic material is typically poor in Norway, dating by means of shoreline displacement is often the only and most precise and accurate method by which one can hope to date the sites. Dating by means of shoreline displacement is consequently fundamental to our understanding of the Norwegian Stone Age. However, this is both because shoreline dating is fundamental to the temporal framework on which our understanding of the period is based, but also because the method is only applicable so long as the societies in question have continuously settled on or close to the contemporary shore-line. Consequently, adherence or deviation from this pattern has major implications for the socio-economic foundations of the societies in question.

@@

Despite the fundamental role of shoreline-dating for Norwegian archaeology, there have only been conducted relatively small scale investigations into the applicability of dating by means of shore-line displacement, and the relationship has only been evaluated using relatively coarse methods. The aim of this paper is to provide a systematic and comprehensive review of the degree to which radiocarbon dates correspond with the dates informed by our current knowledge of shoreline displacement in a larger area of south-eastern Norway, using a more refined methodological approach. The goal is to quantify the degree to which the assumption of shore-bound settlement holds through the Stone Age. This can be used both to inform the certainty with which one can depend on dating by means of shoreline displacement, as well as its precision.  If there is chronological variation in the degree to which the sites were shore-bound, the investigation can also help identify societal shifts, and any such variation can potentially also have implications for how archaeological investigations should be directed at different altitudes in the area of analysis. As presented in more detail below, this problem involves the combined evaluation of three major analytical dimensions. One is the questions of when the sites were in use, the second pertains to the reconstruction of the contemporaneous sea-level, and the third follows from the fact that the relation between site and shore-line is inherently a spatial question. Taking inspiration from recent studies that have integrated multiple sources of spatio-temporal uncertainty, a Monte Carlo simulation-based approach was adopted for the analysis.

The approach is only used to inform the question of the distance between sites and shorelines here, but this method can also be extended to a wide range of use-cases where one needs to evaluate and visualise the relationship between archaeological phenomena, the prehistoric shore-line, and the uncertainty inherent in this reconstruction.

@@

# Methods

The goal of this study is effectively to evaluate the spatial relationship between two phenomena, sites and shore-lines, each associated with their own range of temporal uncertainty. The first task was therefore to ascribe likely date ranges and a numerical expression of the associated uncertainty to the two phenomena.

Due to the dsfg there is a gradient in the rate of isostatic rebound, with increasing rebound rates towards the north-east. To take account of this gradient, the shoreline displacement curves were adjusted to each site location by taking into account the distance to the closest isobase. This was done by first finding the difference between the two closest displacement curves. The elevation difference was found for each year along the entirety of the curves, with each value divided by the meter distance between the isobases. This was done for both ends of the confidence intervals. To find the adjusted displacement curve for a given site location the distance between the site polygon and closest isobase was multiplied with the values for difference per meter. If the site is located south-west of the closest isobase, the elevation values were then subtracted from the original displacement curve, and if the site is located to the north-east the values were instead added to the original curve. The result of this process is shown for an example site in Figure.

For the date ranges associated with the sites, all radiocarbon dates were first individually calibrated using the IntCal20 calibration curve using OxCal v. . All dates not younger than 1700 BCE with 98 % certainty were then excluded from further analysis. Radiocarbon dates associated with each site were then inspected, and with reference to the evaluation done in the original reports, dates believed to belong to the same settlement phase were grouped. In the case were there are multiple dates believed to belong to a single phase, these were subjected to Bayesian modelling using the Boundary function in OxCal. Multiple phases at a single site were treated as independent of each other (Figure x).

As the excavation of archaeological sites typically follow from residential and commercial development and the expansion of infrastructure, the area immediately surrounding the sites have sometimes been severely impacted by modern disturbances. In addition to employing 10 m resolution DTM to alleviate some of these issues, this also necessitated some additional editing of the elevation raster. This involved manually defining the extent of problem areas such as railways, highways, quarries and the like, using polygon features. The DTM values on these were then set to missing, and new elevation values were interpolated from the surrounding terrain. This was done using regularised spline interpolation with tension, using the default settings of r.fillnulls through GRASS GIS. In addition to being available in the , it has also been noted in the supplementary material when the area surrounding a site has been edited in this manner.

Armed with a likely date range for the occupation(s) of each site, the likely trajectory of relative sea-level change at that location, and an DTM edited to remove substantial modern , the simulation was performed. Each simulation run involved first drawing a single year from the posterior density estimate of a given occupation phase. This year then has a corresponding likely elevation range of the shoreline. An elevation value was drawn uniformly from this range, using intervals of 5 cm. The sea-level was then raised to this elevation on the DTM by defining all elevation values at or below this altitude as missing, and creating polygons of the resulting areas with missing values. The horisontal distance was then found by measuring the shortest distance between site and sea polygons, and the vertical distance by subtracting the. The so-called topographic distance between site and sea was also found by measuring the distance while taking into account the slope of the terrain on the DTM. This was done using the topoDist package. In the case where the sea-polygons intersects the site polygon, all distance measures were set to zero. In the case that the sea-polygons completely contain the site, the horisontal and topographic distance measures were made negative. While its safe to assume that an archaeological site was not occupied when it was located beneath sea-level, a negative result can reflect the inherent uncertainty in this procedure and might also help identify discrepancies in displacement data or radiocarbon dates.

horisontal and topographic distance between site and shore-line was measured.  
This process was repeated 1000 times for each phase for each site. The choice of 1000 iterations follows an evaluation of convergence on median distances using the example site Hovland 3. At 1000 simulation runs the analysis of each site ranged from  hours, depending on the distance to the simulated shorelines and consequently the necessary size of the window of analysis.

The distances were then measured.

### 2021-11-01 analysis/supplementary-materials/supplementary.Rmd
More or less set up to run through all analyses (missing Brunstad site limits, and need to enter two Kvastad survey dates, also check Kvastad duplication of dates).
Stepping through each site

Simulation results for each individual site. The first plot on the first row for each site shows all radiocarbon dates from the site. The fill colour of the radiocarbon dates indicates whether they are interpreted as belonging to the same phase and each "sum" indicates that the dates for a phase have been modelled using the Boundary function and then summed. A red outline indicates that the date(s) were not seen as related to the lithic inventory in the original report (see also main text). The second plot on the first row displays the site location in the present day landscape on the edited DTM (i.e. highways and railways have been removed and elevation values are interpolated from the surroundings; see analysis/script/03script.R). The third plot on the first rows shows the site location within the study area relative to the isobases of the displacement curves in use (see main text for more on these).

>>>>

This document includes some notes and shows the simulation results for each individual site analysed in the study. The first plot on the first row for each site shows all calibrated radiocarbon dates from the Stone Age associated with the site if there are more than a single date (see analysis/data/raw_data/radiocarbon.csv for all dates, including those falling outside of the Stone Age). The fill colour of the radiocarbon dates indicates whether they are interpreted as belonging to the same phase, and each sum indicates that the dates for a phase have been modelled using the Boundary function and then summed using OxCal. Each phase is assumed to be independent of other phases. A red outline indicates that the date(s) were not seen as related to the lithic inventory in the original report (see also main text). The second plot on the first row displays the site location in the present day landscape on the edited DTM (i.e. highways and railways impacting the adjusted sea-level have been removed and elevation values are interpolated from the surroundings; see main text and analysis/script/03script.R). The third plot on the first row shows the site location within the study area relative to the isobases of the displacement curves in use.

Subsequent rows for each site then show the simulation results. The first plot on these rows shows the probability density function from which site dates were drawn during simulation. The second plot shows the result of the simulation runs, where the intensity of the colour indicates the number of times the sea was simulated to be present at any given location. The third plot displays a violin plot of the measured distances between site and shore-line across all simulation runs. Measures for central tendency and dispersion associated with these are then given in the table

Gunnarsrød 4 situated right by the railway (see also Gunnarsrød 5 below). The date is seen as possibly the result of old wood effect, a notion that is in line with simulation results here, as this date would the site located beneath the sea-level.
